---
title: "5. Data Structure - Part 2"
description: |
   In this lesson we will review different data structures in R, including data frames and tibbles.
# author:
#   - name: Jenny Sloane
date: 2023-02-16
output:
  distill::distill_article:
    self_contained: false
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

A data structure is a unique way of organizing data so that it can be stored, processed, and retrieved effectively. In the previous Data Structure lesson we reviewed vectors, matrices, and arrays. And now in this lesson we will review:

- Data frames
- Tibbles

## Data frames

What is a data frame?
  - A data frame is a common way of storing data in R. Data frames can have variables are different types (e.g. one variable may be a character and one may be numeric).
  
I find the best way to understand a new concept (especially with coding) is to work through a concrete example. So here, we will explore the penguins dataset that is included in the palmerpenguins package, so once you load the library you can take a look at `penguins`
- how can we interpret this data?
  - each row is a unique penguin
  - each column is a variable that provides information about the penguin 
- we can see this is a data frame and it actually tells us penguins is specifically a "tibble"
- tibbles are simple data frames with some features that make working with the data really nice
  - shows us exactly how big our dataset is (344 rows and 8 columns)
  - automatically only prints out the first 10 rows, so we can always see the variable names at the top
  - under each variable name, we see what type of variable it is (i.e. fct, dbl, int)

```{r eval=FALSE}
penguins

is.data.frame(penguins)
is_tibble(penguins)

is.data.frame(names)
```

## Tibbles


### Create our own tibble

- let's combine the vectors we created in the previous workshop into a tibble using `tibble()`

```{r eval=FALSE}
names <- c("xxx", "yyy", "zzz")
age <- c(20, 18, 33)
blue_eyes <- c(TRUE, FALSE, FALSE)

my_tibble <- tibble(
  Name = names,
  Age = age,
  Blue_eye = blue_eyes
)

is.data.frame(my_tibble)
is_tibble(my_tibble)

my_tibble$Name[1]

my_tibble[1,3]


```


